---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(joker)
```

```{r srr-tags, eval = FALSE, echo = FALSE}
#' @srrstats {G1.2} A Life Cycle Statement describing current and anticipated future states of development is included in the CONTRIBUTING.md file as suggested by the `srr` development team.
#' @srrstats {G1.4, G1.4a} All functions, exported and internal, are documented using [`roxygen2`](https://roxygen2.r-lib.org/)
#' @srrstats {G1.5, G1.6} The package vignette includes all code necessary to reproduce results which form the basis of performance claims made as well as alternative implementations in other R packages.
#' @srrstats {G3.0} All tests compare floating point numbers for approximate equality using tolerance. Please check the test files.
#' @srrstats {G5.1} Only direct simulations (with `set.seed()`) are used in the
#' package. No data sets are created. Users can reproduce these simulations and
#' run examples.
#' @srrstats {G5.2, G5.2a, G5.2b, G5.3} All warnings and errors are covered in the package tests, as verified by `goodpractice::gp()` and covr. The messages produced are fairly descriptive in order to guide the users and the developers. The `check_data()` function checks the data for missing and undefined values, and tests verifying that these values are appropriately handled are included in the test files.
#' @srrstats {G5.10} Extended tests concerning estimator consistency and asymptotic variance are included and run under a common framework with other tests but can be switched on/off by the `JOKER_EXTENDED_TESTS="true"` environment variable.
#' @srrstats {G5.12} The expected runtime of the basic and the extended tests are described in the `tests/README.md` file.
```

# joker <img src=`r file.path("man", "figures", "logo.png")` align="right" height="139" alt="logo"/>

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/joker)](https://CRAN.R-project.org/package=joker)
[![R-CMD-check](https://github.com/thechibo/joker/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/thechibo/joker/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/thechibo/joker/graph/badge.svg)](https://app.codecov.io/gh/thechibo/joker)
[![pkgcheck](https://github.com/thechibo/joker/workflows/pkgcheck/badge.svg)](https://github.com/thechibo/joker/actions?query=workflow%3Apkgcheck)
[![CircleCI build status](https://circleci.com/gh/thechibo/joker.svg?style=svg)](https://circleci.com/gh/thechibo/joker)
<!-- badges: end -->

## Introduction

The `joker` R package develops an S4 distribution system and performs parameter estimation in common distribution families, making well-established and state-of-the-art methods more accessible.

### Key Features

1. The common d, p, q, r function family for each distribution (e.g. dnorm, pnorm, qnorm, rnorm) is enriched with 
  - the ll counterpart (e.g. llnorm) that calculates the log-likelihood, 
  - the e counterpart (e.g. enorm) that performs parameter estimation,
  - the v counterpart (e.g. vnorm) that calculates the asymptotic variance-covariance matrix of an estimator.
2. An S4-class distribution system is developed, allowing the generic evaluation of the dpqr function family and basic distribution calculus. 
3. Moment functions (mean, median, mode, var, sd, skew, kurt) as well as functions that calculate the entropy and the Fisher Information are available for all distributions.
4. Distributions not included in base R are made available, such as the Dirichlet and the Multivariate Gamma.
5. Parameter estimation is performed analytically instead of numerically for the estimators that can be expressed explicitly.
6. Numerical optimization of the MLE (whenever required, e.g. the Beta and Gamma distributions) is performed with computational efficiency, taking advantage of the score equation system to reduce the dimensionality of the optimization.
7 Functions to compute and plot common estimator metrics (bias, variance, and RMSE) are included in the package to allow the convenient study and comparison of the estimators.

## Installation

You can install the release version of `joker` from CRAN by running the following line of code:

```{r, eval=FALSE}
 install.packages("joker")
```

You can install the development version of `joker` from github by running the following line of code:

```{r, eval=FALSE}
 devtools::install_github("thechibo/joker")
```

More details can be found in the [joker Github repository](https://github.com/thechibo/joker "joker Github repository").

## Documentation

Detailed documentation, along with reproducible examples, can be found in the package vignette `vignette(topic = "joker", package = "joker")`.

## Team

The `joker` package is developed in the 
[Mathematics Department](https://en.math.uoa.gr/ "Mathematics Department Homepage") of the [University of Athens](https://en.uoa.gr/ "University of Athens Homepage"). The package maintainer is [Ioannis Oikonomidis](http://users.uoa.gr/~goikon/ "Ioannis Oikonomidis Homepage"), working under the supervision of [Prof. Samis Trevezas](http://scholar.uoa.gr/strevezas/ "Samis Trevezas Homepage").
